\tit

### 2021-06-03 paper.tex
misc near-final tweaks

This paper makes the following specific contributions:
\begin{itemize}
\item
  A collection of lemmas are stated and proved about functions over lists, capturing the semantic essence of Brzozowski's syntactic technique of derivatives of regular expressions (\secref{Decomposing Languages}).
\item
  A duality is revealed between regular expressions and tries (prefix trees) in the context of language differentiation (\secref{From Languages to Parsing}).
  Each implementation contains exactly the same code but in transposed forms.
  Both are corollaries of the semantic lemmas and are automatically proved correct by type-checking.
\item
  The theory and algorithms are specified in terms of proof isomorphism rather than the usual, coarser relation of logical equivalence.
  As such, the lemmas for language concatenation and Kleene closure are more informative than in previous work, capturing \emph{all} parsings distinctly, including closure of languages containing the empty string.
\item
  The vocabulary of languages is generalized from predicates on lists to predicates on arbitrary types, and then re-specialized to the usual vocabulary (\secref{Predicate Algebra}).
  This generalization consists of two useful covariant applicative functors on functions (the usual one on functions-from-\AF{X} and another on functions-to-\APT{Set}).
  % Properties are proved in terms of this more general vocabulary.
\item
  Properties are stated first at the level of predicates, making them easy to understand and prove, thanks to elimination of operational details.
  These properties are then transported to the computable parsing implementations (\secref{Transporting Properties from Specification to Implementations}) with no additional proofs needed.
\end{itemize}
Although many proofs are elided below, the source code for this paper is freely available and contains fully verified Agda code with no postulates.%
\ifacm\else
%% \citep{Elliott2021-derivatives-repo}.
%% (\repoURL).
\footnote{\protect\repoURL}
\fi{}
All type signatures and definitions in the paper are generated by the Agda compiler.

@@

The ``symbolic differentiation'' style reflects language-building vocabulary (\figref{Lang-ops}) into an inductive data type (of modestly extended regular expressions), while \AF{ν} and \AF{δ} become functions defined by pattern-matching on that data type.
For convenience, use the same names as in \secref{Specifying Languages} for these new counterparts, while referring to the original versions via the module prefix ``{◇.}\hspace{0.05em}''.
The result is shown in \figrefdef{symbolic-api}{Regular expressions (inductive)}{\api} along with a trivially defined semantic function {\AF{⟦\_⟧}} (mapping any representation of type {\AF{Lang} \AB P} to the corresponding type-level language \AB{P}) and a less trivial, computable function  {\AF{⟦\_⟧‽}} that converts a symbolic language representation to a computable parser.
Note that the \AF{\_◂\_} operation from \secref{Decidability} has been reified as a new constructor alongside the more conventional language operators.
The reason \AF{\_◂\_} must be part of the inductive representation is the same as the other constructors, namely so that the core lemmas (\figref{nu-delta-lemmas}) translate into an implementation in terms of that representation.

Decidable parsing relies only on \AF{ν} and \AF{δ} (via {\AF{⟦\_⟧‽}}), which are defined in \figrefdef{symbolic-defs}{Symbolic differentiation (column-major/patterns)}{\agda{defs}}, as systematically derived from the lemmas of \figref{nu-delta-lemmas}.\out{\notefoot{Describe this derivation.}}
These clauses are meant to be read in ``column-major'' order, i.e., each column is a complete definition.
Correctness is guaranteed by the types of \AF{ν} and \AF{δ} and so is proved automatically by type-checking \figref{symbolic-defs}.
(Recall from \secref{Decidability} that correctness is defined as constructing a term of type {\AF{Decidable} \AF{P}} for a given language \AB{P}, as satisfied by the type of {\AF{⟦\_⟧‽}}.)
We could thus use \emph{any} definitions of \AF{ν} and \AF{δ} that type- and termination-check, although not many definitions would do so.

@@

Decidable recognition again relies only on \AF{ν} and \AF{δ}, which are defined for each language operation in \figrefdef{automatic-defs}{Automatic differentiation (row-major/copatterns)}{\agda{defs}} (though with a problem to be noted and fixed in the next paragraph).
These clauses are meant to be read in ``row-major'' order, i.e., each row is a complete definition.
Note that the clauses in \figref{automatic-defs} are syntactically identical to those in \figref{symbolic-defs} but are organized dually.
(The compiler-generated syntax coloring differs to reflect the changed interpretation of the definitions.)

>>>>

This paper makes the following specific contributions:
\begin{itemize}
\item
  A collection of lemmas are stated and proved about functions over lists, capturing the semantic essence of Brzozowski's syntactic technique of derivatives of regular expressions (\secref{Decomposing Languages}).
\item
  A duality is revealed between regular expressions and tries (prefix trees) in the context of language differentiation (\secref{From Languages to Parsing}).
  Each implementation contains exactly the same code but in transposed forms.
  Both are corollaries of the semantic lemmas and are automatically proved correct by type-checking.
\item
  The theory and algorithms are specified in terms of proof isomorphism rather than the usual, coarser relation of logical equivalence.
  As such, the lemmas for language concatenation and Kleene closure are more informative than in previous work, capturing \emph{all} parsings distinctly, including closure of languages containing the empty string.
\item
  The vocabulary of languages is generalized from predicates on lists to predicates on arbitrary types, and then re-specialized to the usual vocabulary (\secref{Predicate Algebra}).
  This generalization consists of two useful covariant applicative functors on functions (the usual one on functions-from-\AF{X} and another on functions-to-\APT{Set}).
  % Properties are proved in terms of this more general vocabulary.
\item
  Properties are stated first at the level of predicates, making them easy to understand and prove, thanks to elimination of operational details.
  These properties are then transported to the computable parsing implementations (\secref{Transporting Properties from Specification to Implementations}) with no additional proofs needed.
\end{itemize}
Although many proofs are elided below, the source code for this paper is freely available\footnote{\protect{\repoURL}} and contains fully verified Agda code with no postulates.
The type signatures and definitions displayed in the paper are extracted from this source code by the Agda compiler.

@@

The ``symbolic differentiation'' style reflects language-building vocabulary (\figref{Lang-ops}) into an inductive data type (of modestly extended regular expressions), while \AF{ν} and \AF{δ} become functions defined by pattern-matching on that data type.
For convenience, use the same names as in \secref{Specifying Languages} for these new counterparts, and refer to the original versions via the module prefix ``{◇.}\hspace{0.05em}''.
The result is shown in \figrefdef{symbolic-api}{Regular expressions (inductive)}{\api} along with a computable function {\AF{⟦\_⟧‽}} that converts a symbolic language representation to a computable parser.
Note that the \AF{\_◂\_} operation from \secref{Decidability} has been reified as a new constructor alongside the more conventional language operators.
The reason \AF{\_◂\_} must be part of the inductive representation is the same as the other constructors, namely so that the core lemmas (\figref{nu-delta-lemmas}) translate into an implementation in terms of that representation.

The meaning of any {\AB r \AK{:} \AF{Lang} \AB P} is defined simply as \AB{P}, ignoring the representation \AB{r} itself\out{ (beyond its type)}:
\agda{semantics}

Decidable parsing relies only on \AF{ν} and \AF{δ} (via {\AF{⟦\_⟧‽}}), which are defined in \figrefdef{symbolic-defs}{Symbolic differentiation (column-major/patterns)}{\agda{defs}}, as systematically derived from the lemmas of \figref{nu-delta-lemmas}.\out{\notefoot{Describe this derivation.}}
These clauses are meant to be read in column-major order, i.e., each column is a complete definition.
Correctness is guaranteed by the types of \AF{ν} and \AF{δ} and so is proved automatically by type-checking \figref{symbolic-defs}.
(Recall from \secref{Decidability} that correctness is defined as constructing a term of type {\AF{Decidable} \AF{P}} for a given language \AB{P}, as satisfied by the type of {\AF{⟦\_⟧‽}}.)
We could thus use \emph{any} definitions of \AF{ν} and \AF{δ} that type- and termination-check, although not many definitions would do so.

@@

Decidable recognition again relies only on \AF{ν} and \AF{δ}, which are defined for each language operation in \figrefdef{automatic-defs}{Automatic differentiation (row-major/copatterns)}{\agda{defs}} (though with a problem to be noted and fixed in the next paragraph).
These clauses are meant to be read in row-major order, i.e., each row is a complete definition.
Note that the clauses in \figref{automatic-defs} are syntactically identical to those in \figref{symbolic-defs} but are organized dually.
(The compiler-generated syntax coloring differs to reflect the changed interpretation of the definitions.)
